---
title: "Figure-it-out-markdown"
author: "Trevor Kapuvari & Timothy Oliver"
date: "15 December 2023"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---


```{r clear_environment, include=F, results = 'hide'}
rm(list=ls())
```


```{r setup, include=FALSE}

# global options for knitting chunks
local({
  hook_output <- knitr::knit_hooks$get('output')
  knitr::knit_hooks$set(output = function(x, options) {
    if (!is.null(options$max.height)) options$attr.output <- c(
      options$attr.output,
      sprintf('style="max-height: %s;"', options$max.height)
    )
    hook_output(x, options)
  })
})
knitr::opts_chunk$set(echo = TRUE,message=F,warning=F)



library(summarytools)
library(stargazer)
library(vtable)
library(tidyverse)
library(tidycensus)
library(sf)
library(spdep)
library(caret)
library(ckanr)
library(FNN)
library(grid)
library(gridExtra)
library(ggcorrplot) # plot correlation plot
library(corrr)      # another way to plot correlation plot
library(kableExtra)
library(jtools)     # for regression model plots
library(ggstance) # to support jtools plots
library(ggpubr)    # plotting R^2 value on ggplot point scatter
library(broom.mixed) # needed for effects plots
library(tufte)
library(viridis)
library(RSocrata)
library(classInt)
library(spatstat)
library(pscl)
library(plotROC)
library(pROC)
library(lubridate)
library(ROSE)
library(tigris)
library(knitr)
library(RSocrata)
library(gganimate)


# functions and data directory
root.dir = "https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/DATA/"


source("https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/functions.r")
source("themes/plotTheme_TO.r")
# source("functions/quantileBreaks.r")

palette2 <- c("#981FAC","#FF006A")
palette5 <- c("#25CB10", "#5AB60C", "#8FA108",   "#C48C04", "#FA7800")
palette5a <- c("#eff3ff","#bdd7e7","#6baed6","#3182bd","#08519c")
palette4a <- c("#D2FBD4","#92BCAB","#527D82","#123F5A")
palette2a <- c("#6baed6","#08519c")

palette9 <- viridis(9)

options(scipen=999)

```

# Here We Go

To download the data, go [here](https://www.kaggle.com/datasets/pranavbadami/nj-transit-amtrak-nec-performance/) using the button in the top-right and place the zip file in the data folder. The `kaggle_data` chunk should handle the rest.

```{r kaggle_data, results='hide'}
if (!dir.exists("data/Kaggle_transit&amtrak_data/")){
unzip("data/Kaggle_transit&amtrak_data.zip",exdir="data/Kaggle_transit&amtrak_data")
}

datApr2020 <- read.csv("data/Kaggle_transit&amtrak_data/2020_04.csv")
datMay2020 <- read.csv("data/Kaggle_transit&amtrak_data/2020_05.csv")


datApr2020 <- datApr2020 %>%
  mutate(interval60 = floor_date(ymd_hms(scheduled_time), unit = "hour"),
         interval15 = floor_date(ymd_hms(scheduled_time), unit = "15 mins"),
         interval1 = floor_date(ymd_hms(scheduled_time), unit = "min"),
         week = week(interval60),
         dotw = wday(interval60,label=T),
         delayed = ifelse(delay_minutes > 0,TRUE,FALSE))

datMay2020 <- datMay2020 %>%
  mutate(interval60 = floor_date(ymd_hms(scheduled_time), unit = "hour"),
         interval15 = floor_date(ymd_hms(scheduled_time), unit = "15 mins"),
         interval1 = floor_date(ymd_hms(scheduled_time), unit = "min"),
         week = week(interval60),
         dotw = wday(interval60,label=T),
         delayed = ifelse(delay_minutes > 0,TRUE,FALSE))

nj_trans_stations <- st_read("data/Rail_Stations_of_NJ_Transit.geojson")%>% st_transform(crs="ESRI:102711")

nj_trans_lines <- st_read("data/Rail_Lines_of_NJ_Transit.geojson")%>% st_transform(crs="ESRI:102711")

## !!!
### WE LOSE 27 STATIONS BY MERGING THE NAMES TO THE POINT DATA
## !!!

may_w_stations <- merge(x= nj_trans_stations,y=datMay2020,by.x="STATION_ID",by.y="from")%>%
  rename(from_station = STATION_ID,
         fromLat = LATITUDE, fromLon = LONGITUDE, 
         from_atis_id = ATIS_ID, from_muni = MUNICIPALITY,
         from_rail_service = RAIL_SERVICE)%>%
  select(-OBJECTID_1,-LINE_CODE)%>%
  st_drop_geometry()%>%
  merge(y=nj_trans_stations,by.x="to",by.y="STATION_ID")%>%
  rename(to_station = to,
         toLat = LATITUDE, toLon = LONGITUDE, 
         to_atis_id = ATIS_ID, to_muni = MUNICIPALITY,
         to_rail_service = RAIL_SERVICE) %>%
  select(-OBJECTID_1)%>%
  st_drop_geometry() %>%
  merge(y=nj_trans_lines,by="LINE_CODE")%>%
  # rename(toLat = LATITUDE, toLon = LONGITUDE, 
  #        to_atis_id = ATIS_ID, to_muni = MUNICIPALITY,
  #        to_rail_service = RAIL_SERVICE,
  #        to_line_code = LINE_CODE) %>%
  select(-OBJECTID)%>%
  st_drop_geometry()
```

  
  
```{r simple_map, results='hide'}
census <- 
  get_acs(geography = "county", 
          variables = c("B01003_001", "B19013_001", 
                        "B02001_002", "B08013_001",
                        "B08012_001", "B08301_001", 
                        "B08301_010", "B01002_001"), 
          year = 2021, 
          geometry = TRUE, 
          output = "wide") %>%
  rename(Total_Pop =  B01003_001E,
         Med_Inc = B19013_001E,
         Med_Age = B01002_001E,
         White_Pop = B02001_002E,
         Travel_Time = B08013_001E,
         Num_Commuters = B08012_001E,
         Means_of_Transport = B08301_001E,
         Total_Public_Trans = B08301_010E)%>%
  st_transform(crs="EPSG:4326")

tracts <- 
  get_acs(geography = "tract", 
          variables = c("B01003_001", "B19013_001", 
                        "B02001_002", "B08013_001",
                        "B08012_001", "B08301_001", 
                        "B08301_010", "B01002_001"), 
          year = 2021, 
          geometry = TRUE,
          state="NJ",
          output = "wide") %>%
  rename(Total_Pop =  B01003_001E,
         Med_Inc = B19013_001E,
         Med_Age = B01002_001E,
         White_Pop = B02001_002E,
         Travel_Time = B08013_001E,
         Num_Commuters = B08012_001E,
         Means_of_Transport = B08301_001E,
         Total_Public_Trans = B08301_010E)%>%
  st_transform(crs="EPSG:4326")

threeStates = census%>%filter(grepl("New Jersey",NAME)|grepl("Pennsylvania",NAME)|grepl("New York",NAME))
nj = census%>%filter(grepl("New Jersey",NAME))
```


## Figure this Out
```{r census_join}
may_census <- st_join(may_w_stations %>% 
          filter(is.na(fromLon) == FALSE &
                   is.na(fromLat) == FALSE &
                   is.na(toLon) == FALSE &
                   is.na(toLat) == FALSE) %>%
          st_as_sf(., coords = c("fromLon", "fromLat"), crs = 4326),
        threeStates %>%
          st_transform(crs=4326),
        join=st_intersects) %>%
  mutate(fromCountyLon = unlist(map(geometry, 1)),
         fromCountyLat = unlist(map(geometry, 2)))%>%
  as.data.frame() %>%
  rename(Origin.County = GEOID) %>%
  select(-geometry)%>%
  st_as_sf(., coords = c("toLon", "toLat"), crs = 4326) %>%
  st_join(., threeStates %>%
            st_transform(crs=4326),
          join=st_intersects) %>%
  rename(Destination.County = GEOID)  %>%
  mutate(toCountyLon = unlist(map(geometry, 1)),
         toCountyLat = unlist(map(geometry, 2)))%>%
  as.data.frame() %>%
  select(-geometry)
```

```{r view_stations}

ggplot()+
  # geom_sf(data=threeStates%>%st_transform(st_crs(nj_trans_stations)),color = "white",fill=GRAY9)+
  geom_sf(data=nj%>%st_transform(st_crs(nj_trans_stations)),color = "white",fill=GRAY9)+
  geom_sf(data=nj_trans_stations,color="blue")+
  geom_sf(data=nj_trans_lines,color = GRAY4)+
  plotTheme()+
  theme(panel.background=element_rect(
        fill = "#fcfcf4",
        colour = NA),
        panel.grid.major=element_blank(),
        axis.text=element_blank())
```

## Exploratory Analysis

```{r plots}

expA = may_w_stations[is.na(may_w_stations$delayed)==T,]$delayed="Amtrak"

ggplot(may_w_stations %>%
         group_by(interval60) %>%
         tally())+
  geom_line(aes(x = interval60, y = n))+
  labs(title="Train Trips in NYC Area, May, 2020",
       x="Date", 
       y="Number of trips")+
  plotTheme()+
  coord_cartesian(ylim=c(0,300))

ggplot(may_w_stations,mapping=aes(x=delayed))+
  geom_bar(aes(fill=delayed,show.legend=F))+
  geom_text(aes(label=..count..),stat="count",color="white",vjust=1.5)+
  labs(title="Delayed Train Trip Density in NYC Area, May, 2020",
       y="Number of trips",
       x = "Delayed")+
  guides(fill=guide_legend(title = "Delayed Train?"))+
  scale_color_manual(labels=c("False","True","Amtrak"),
                     values = c(GRAY9,BLUE3,GREEN3) # did nothing?
                     )+
  plotTheme()+
  theme(panel.border = element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.y = element_blank())

# # forgot what this plot was going to be
# ggplot(expA %>%
#          group_by)+
#   geom_bar(aes(x=delayed,fill=delayed,show.legend=F))+
#   geom_text(aes(label=..count..),stat="count",color="brown",vjust=1.5)+
#   labs(title="Delayed Train Trip Density in NYC Area, May, 2020",
#        y="Number of trips")+
#   plotTheme()

```

**Note: All `r length(datMay2020[datMay2020$type=="Amtrak",]$interval60)` of Amtrak's records in the full May data have no scheduled time and thereby no interval. The merged data has `r length(may_w_stations[may_w_stations$type=="Amtrak",]$interval60)` such NA records**


```{r from_station}
ggplot(may_w_stations %>%
         group_by(interval60, from_station) %>%
         tally())+
  geom_histogram(aes(n), binwidth = 1, color = GRAY8)+
  labs(title="Train trips per hr by station. NJTransit & Amtrak Area, May, 2020",
       x="Trip Counts", 
       y="Number of Stations")+
  plotTheme()+
  coord_cartesian(xlim=c(0,10))
```


### panel 
```{r panel}
length(unique(may_w_stations$interval60)) * length(unique(may_w_stations$from_id))


study.panel <- 
  expand.grid(interval60=unique(may_w_stations$interval60), 
              from_id = unique(may_w_stations$from_id)) %>%
  left_join(., may_census %>%
              # changed fromLon & Lat to fromCountyLon & Lat
              select(train_id, from_id, from_station, fromCountyLon, fromCountyLat,line,delay_minutes)%>% # Origin.Tract,
              distinct() %>%
              group_by(from_id) %>%
              slice(1))

nrow(study.panel)
```


```{r geometry_in_panel}
ride.panel <- may_census %>%
  mutate(Trip_Counter = 1) %>%
  right_join(study.panel) %>% 
  # changed fromLon & Lat to fromCountyLon & Lat
  group_by(interval60, from_id, from_station, Origin.County, fromCountyLon, fromCountyLat,delay_minutes) %>%
  summarize(Trip_Count = sum(Trip_Counter, na.rm=T)) %>%
  ungroup() %>%
  filter(is.na(from_id) == FALSE) %>%
  mutate(week = week(interval60),
         dotw = wday(interval60, label = TRUE)) %>%
  filter(is.na(Origin.County) == FALSE) %>%
  # left_join(study.panel, threeStates %>%
  #             as.data.frame()) %>% # extra parenthesis
              # select(-geometry), by = c("Origin.Tract" = "GEOID")) %>%
  arrange(from_id, interval60) %>% 
  mutate(lagHour = dplyr::lag(delay_minutes,1),
         lag2Hours = dplyr::lag(delay_minutes,2),
         lag3Hours = dplyr::lag(delay_minutes,3),
         lag4Hours = dplyr::lag(delay_minutes,4),
         lag12Hours = dplyr::lag(delay_minutes,12),
         lag1day = dplyr::lag(delay_minutes,24)) %>%
         # # Indigenous Peoples' (Columbus) Day is federally recognized
         # #  but not an official city holiday of Austin
         # ## I include it to account for holiday lag
         # holiday = ifelse(yday(interval60) == 282,1,0)) %>%
   mutate(day = yday(interval60)) # %>%
   # mutate(holidayLag = case_when(dplyr::lag(holiday, 1) == 1 ~ "PlusOneDay",
   #                               dplyr::lag(holiday, 2) == 1 ~ "PlusTwoDays",
   #                               dplyr::lag(holiday, 3) == 1 ~ "PlusThreeDays",
   #                               dplyr::lead(holiday, 1) == 1 ~ "MinusOneDay",
   #                               dplyr::lead(holiday, 2) == 1 ~ "MinusTwoDays",
   #                               dplyr::lead(holiday, 3) == 1 ~ "MinusThreeDays"),
   #       holidayLag = ifelse(is.na(holidayLag) == TRUE, 0, holidayLag))

```

### get lines and visualize reliability


### timelapse of reliability (week)

```{r timelapse_beg}
may_census %>% 
  merge(y=census,by.x="Origin.County",by.y="GEOID") %>%
  group_by(interval60, from_station, geometry,Origin.County) %>% 
  tally() %>% 
  rename(departures = n) %>%
  st_as_sf() %>%
  st_transform(crs = st_crs(nj_trans_stations))

plotCheckoutData <- may_census[yday(may_census$interval60)==138,] %>% 
  merge(y=census,by.x="Origin.County",by.y="GEOID") %>% 
  group_by(interval60, from_station, geometry,Origin.County) %>% 
  tally() %>% 
  rename(departures = n) %>%
  st_as_sf() %>%
  st_transform(crs = st_crs(nj_trans_stations))
  
plotReturnData <- may_census[yday(may_census$interval60)==138,] %>% 
  merge(y=census,by.x="Destination.County",by.y="GEOID") %>% 
  group_by(interval60, to_station, geometry,Destination.County) %>% 
  tally() %>% 
  rename(arrivals = n) %>%
  st_as_sf() %>%
  st_transform(crs = st_crs(nj_trans_stations))
```


```{r depart_animation}
ggplot()+
  geom_sf(data = plotCheckoutData,color="white",fill="beige")+
  transition_null()+
  geom_sf(data = plotCheckoutData,color="white",aes(fill = departures))+
  # geom_sf(data = cPoints[cPoints$interval60 %in% plotCheckoutData$interval60,],color=GRAY9)+
  plotTheme()+
  # theme(panel.background = "beige",
        # axis.text = element_blank())+
  scale_fill_viridis(direction=-1,option="C",discrete=F)+
  labs(title="Train Trips by Origin County, May, 2020")+
  transition_manual(interval60)
  # coord_sf(crs = st_crs(plotCheckoutData), datum = NA)



```

```{r return_animation}

ggplot()+
  geom_sf(data = plotReturnData,color="white",fill="beige")+
  transition_null()+
  geom_sf(data = plotReturnData,color="white",aes(fill = returns))+
  # geom_sf(data = rPoints[rPoints$interval60 %in% plotReturnData$interval60,],color=GRAY9)+
  plotTheme()+
   # theme(panel.background = "beige",
        # axis.text = element_blank())+
  scale_fill_viridis(direction=-1,option="C",discrete=F)+
  labs(title="Train Trips by Destination County, May, 2020")+
  transition_manual(interval60)

```

### add in the envelope of time delay (sd in prediction)
